Nix for Polyglot, Reproducible Data Science Workflows

### 2025-10-23 paper.qmd
link to pdf
container -> docker container

Once the image is built, the analysis can be executed inside a *container*,
which is a running instance of the image. Containers are typically run
non-interactively, allowing the complete computational environment to be
instantiated and executed with a single command.

@@

1. Poor interactive support: While containers can be modified at runtime,
  changes are lost when the container stops unless explicitly saved. Running
  graphical applications like RStudio Desktop from containers requires complex
  X11 configuration that works reliably only on Linux. Instead, web-based IDEs
  (like RStudio Server) are used, but come with the inherent limitations of
  web-based solutions.

@@

5. Incomplete Reproducibility: Reproducibility should be understood as a
  spectrum rather than a binary property, a notion emphasised in
  "Reproducibility in Software Engineering" by @dellaieraMasterThesis2024.
  Docker illustrates this spectrum well: it provides strong run-time
  reproducibility—that is, consistency in space (where the same image behaves
  identically across different systems) but weak build-time reproducibility,
  which concerns consistency over time. As defined in the thesis,
  reproducibility at build time means being able to produce an identical build
  artefact regardless of when or where it is built. However, Docker inherently
  fails to guarantee this property. Its reliance on mutable base images and
  non-deterministic instructions such as `apt-get update` leads to temporal
  drift: rebuilding the same `Dockerfile` at different points in time can
  produce distinct outputs. In the experimental evaluation conducted in the
  thesis, even with pinned base images, checksums of the resulting containers
  varied across builds and machines. Thus, while Docker offers partial
  reproducibility: strong at run time (space) and weak at build time (time), it
  cannot on its own ensure verifiable, deterministic software artifacts.
  Achieving true reproducibility requires pinning image digests, freezing
  dependency versions, and employing deterministic build systems such as Nix or
  Guix for complete control over both space and time [@Malka_2024].

@@

The `default.nix` file defines the exact computational environment required to
compile the manuscript, ensuring full reproducibility of the build process. To
guarantee the reproducibility of the manuscript, it is automatically
recompiled via GitHub Actions upon each commit, using the same Nix environment.
An HTML version is additionally deployed to GitHub Pages, providing an
accessible format for viewing on smaller devices at this
[link](https://b-rodrigues.github.io/rix_paper/)^[[https://b-rodrigues.github.io/rix_paper/](https://b-rodrigues.github.io/rix_paper/)].

>>>>

Once the image is built, the analysis can be executed inside a *container*,
which is a running instance of the image. [Docker]{.pkg} containers are typically run
non-interactively, allowing the complete computational environment to be
instantiated and executed with a single command.

@@

1. Poor interactive support: While [Docker]{.pkg} containers can be modified at runtime,
  changes are lost when the container stops unless explicitly saved. Running
  graphical applications like RStudio Desktop from containers requires complex
  X11 configuration that works reliably only on Linux. Instead, web-based IDEs
  (like RStudio Server) are used, but come with the inherent limitations of
  web-based solutions.

@@

5. Incomplete Reproducibility: Reproducibility should be understood as a
  spectrum rather than a binary property, a notion emphasised in
  "Reproducibility in Software Engineering" by @dellaieraMasterThesis2024.
  Docker illustrates this spectrum well: it provides strong run-time
  reproducibility—that is, consistency in space (where the same image behaves
  identically across different systems) but weak build-time reproducibility,
  which concerns consistency over time. As defined in the thesis,
  reproducibility at build time means being able to produce an identical build
  artefact regardless of when or where it is built. However, Docker inherently
  fails to guarantee this property. Its reliance on mutable base images and
  non-deterministic instructions such as `apt-get update` leads to temporal
  drift: rebuilding the same `Dockerfile` at different points in time can
  produce distinct outputs. In the experimental evaluation conducted in the
  thesis, even with pinned base images, checksums of the resulting [Docker]{.pkg}
  containers varied across builds and machines. Thus, while Docker offers
  partial reproducibility: strong at run time (space) and weak at build time
  (time), it cannot on its own ensure verifiable, deterministic software
  artifacts. Achieving true reproducibility requires pinning image digests,
  freezing dependency versions, and employing deterministic build systems such
  as Nix or Guix for complete control over both space and time [@Malka_2024].

@@

The `default.nix` file defines the exact computational environment required to
compile the manuscript, ensuring full reproducibility of the build process. To
guarantee the reproducibility of the manuscript, it is automatically
recompiled via GitHub Actions upon each commit, using the same Nix environment.
An HTML version is additionally deployed to GitHub Pages, providing an
accessible format for viewing on smaller devices at this
[link](https://b-rodrigues.github.io/rix_paper/)^[[https://b-rodrigues.github.io/rix_paper/](https://b-rodrigues.github.io/rix_paper/)].
The PDF version of the manuscript can be found at
[link](https://b-rodrigues.github.io/rix_paper/paper.pdf)^[[https://b-rodrigues.github.io/rix_paper/paper.pdf](https://b-rodrigues.github.io/rix_paper/paper.pdf)].
